% ************************** Thesis Acknowledgements **************************

\begin{acknowledgements}      


I owe thanks to many people for helping me to this point. I would like to express my gratitude to my supervisors; Robin Henderson, Jian Qing Shi, and Zhenhong Li.
All of your guidance, support, and knowledge throughout  has helped me considerably. Zhenhong; thank you for introducing me to the world of remote sensing. Jian; many thanks for your immeasurable support even through challenging time differences. Robin; special thanks for stepping in to supervise for the last couple of years, you made what could be a challenging situation a smooth process.

I would like to express my thanks to the Centre for Doctoral Training in Cloud Computing for Big Data for their support. I had the pleasure to work with many brilliant teaching and administrative colleagues, and fellow PhD students. Thanks for your unwavering commitment to research, interesting talks, and friendships.
Thanks should also go to the Engineering and Physical Sciences Research Council for my funding, [grant number EP/L015358/1].

To my family. Thank you for your support through the years of this work; and for all the years that came before it.
In particular, mum and dad, your kindness and love has made this possible.

To Hannah, thank you for being there with me; every step of the way. For putting up with the ups and downs of the whole process. At times when I struggled you were there to help me. I could not have undertaken this journey without you and I am deeply grateful for everything you have done.
\end{acknowledgements}
